\chapter*{Acknowledgment}
\addcontentsline{toc}{chapter}{Acknowledgement}
\begin{comment}

資金や，計算資源，データの提供以外にもこの修士論文を完成させるにあたり，多くの方々のご指導と支援を賜りました．心から感謝の意を表します．

まず，指導教官であるToyotaro Suzumura教授には，研究のテーマ決定，学会選定，スケジュール管理に至るまで，多岐にわたるご指導を賜りました．日々の議論等において、研究の方向性から、人に伝えるためのストーリーの組み立て方など、たくさんのアドバイスをいただきましたが、どれもありがたく、納得のいくものばかりで、より良い研究へと導いてくださいました。また，研究室の雰囲気を良くするための工夫や，締め切り直前にもかかわらず論文の添削をしていただき，深く感謝しております．

次に，同じ研究室のResearch ScientistであるDr. Hiroki Kanezashiには，計算機環境の整備，学会での発表支援，論文の添削, 出張に関する手続き，など，数えきれないほどのご支援をいただき，心より感謝申し上げます．実験においてもアドバイスを多くいただき、実験を効率化できる実践的なtipsを教えていただいたり、結果が芳しく無い時には改善案の相談に乗っていただき、研究の進行に大変助かりました．

また，同じ研究室のAssistant ProfessorであるDr. Masatoshi Hanaiには，学部4年生の卒業研究からお世話になっており、修士課程でも引き続き日常的に、研究に関してお話してくださり、貴重なアドバイスや研究のヒントをいただきました．また，キャリアに関する相談や、研究室での息抜きの雑談、そして研究環境の整備など、日々の研究室生活をより良いものにしていただきました。多方面でのご支援に深く感謝しております．

さらに，同じ研究室をご卒業された先輩であるSatoshi Wakiさんには，修士1年生の時にメンターとして、研究テーマや方向性の決定にたくさん相談に乗ってくださり，多くのアイデアや助言をいただきました．また、議論にもとことん付き合ってくださり、研究を進める大きな推進力となりました。加えて、研究生活やキャリアに関する話もたくさんしていただき、大変ためになりました．誠にありがとうございました．

最後に，いつも支えてくれていた家族に，心からの感謝を表します．皆様の支えがあったからこそ，この論文を完成させることができました．

改めて，これまでお世話になった全ての皆様方に，心から感謝申し上げます．誠にありがとうございました．


\end{comment}
This work was supported by JSPS KAKENHI Grant
JP23H03408, JP21K17749 and JP23K28098.

In addition to funding, % computational resources, and data provision, 
the completion of this master's thesis was made possible through the guidance and support from many individuals, to whom I would like to express my deepest gratitude.

First and foremost, I would like to thank Professor Toyotaro Suzumura, my academic advisor, for his extensive guidance and support throughout my research. He provided me with invaluable advice and direction in deciding on a research theme, selecting conferences and managing my schedule. His insights and suggestions on how to present my research in a clear and compelling manner were particularly helpful. I also appreciate his efforts to foster a positive atmosphere in the lab and a productive research environment, and his willingness to review my papers despite his busy schedule. % Thanks to his support and guidance, my research life has been enriched and fulfilling.

I would also like to thank Dr. Hiroki Kanezashi, a Research Scientist in our laboratory, for his numerous contributions to my research. He provided me with technical support, helped me with presentations, and reviewed my papers. His practical tips on how to conduct experiments efficiently were especially helpful, and his advice on how to improve my results was invaluable. I appreciate his kindness and willingness to help me with various tasks, from setting up computational environments to handling travel arrangements.

I am also profoundly grateful to Dr. Masatoshi Hanai, an Assistant Professor in our laboratory, for his guidance and support.
% I was fortunate to have him as my mentor during my undergraduate days, and even after that, he has continued to be a source of guidance and support for me, offering advice and encouragement whenever I need it. 
His suggestions on how to approach my research and his valuable insights on the research topics were particularly helpful. I also appreciate his comprehensive support, from helping me set up the research environment and equipments, to engaging in daily conversations that were both enjoyable and informative, which has greatly enhanced my research experience.
% his support has not only made my research more efficient, but also more enjoyable and fulfilling.

Furthermore, I would like to thank Satoshi Waki, a senior alumnus of our laboratory, for his mentorship and guidance. He provided me with valuable advice on how to choose a research topic and how to find directions, and his insights on how to conduct research were particularly helpful. His willingness to engage in discussions and provide feedback on my ideas was invaluable, and his advice on the research life and the future career goals was especially helpful.

Last but certainly not least, I would like to acknowledge the unwavering support and encouragement of my family, without whom I would not have been able to complete my research.

Once again, I would like to express my heartfelt gratitude to everyone who has supported me throughout my research life. I am deeply grateful for your kindness, guidance, and encouragement, and I am honored to have had the opportunity to work with such a wonderful group of people. Thank you.
